---
title: "refs check"
format: 
  html:
    embed-resources: true
---

## Range shift reference lists

Compare references from Lenoir et al. 2020 and Rubenstein et al. 2023.

```{r}
library(tidyverse)
library(here)
library(DBI)
```

### Refs from Rubenstein et al. 2023

To match Lenoir et al. 2020, just grab first author's last name...

```{r}
db_file <- here('_raw_data/db_range_shifts.sqlite')
mydb <- dbConnect(RSQLite::SQLite(), db_file)

DBI::dbListTables(mydb)
refs_rubenstein <- dbGetQuery(mydb, 'SELECT * FROM paper_id') %>%
  mutate(first_author = str_remove(author, ' et al.*| and .+')) %>%
  mutate(rubenstein = TRUE)

dbDisconnect(mydb)
```

### Refs from Lenoir et al. 2020

These are listed as just a list of refs and ID with no column names... the details need to be parsed out!  I started to do it with text processing but ended up just dropping it into Gemini and asked it to parse for me.  That result is saved as `_raw_data/lenoir_refs.csv`.

```{r}
refs_lenoir <- read_csv(here('_raw_data/lenoir_refs.csv')) %>%
  mutate(lenoir = TRUE)
# head(refs_lenoir_raw)
```

### try to match

```{r}
author_clean <- function(a) {
  a %>%
    stringi::stri_trans_general('Latin-ASCII')
}
title_clean <- function(t) {
  t_out <- str_to_lower(t) %>% 
    str_replace_all('-', ' ') %>%
    str_remove_all('[^a-z ]') %>%
    str_squish() %>%
    ### fix some typos
    str_replace('chacteristics', 'characteristics') %>%
    str_replace('euroatlantic', 'euro atlantic') %>%
    str_replace('vegetation of mauna', 'vegetation on mauna') %>%
    str_replace('the geographic range', 'the geographical range') %>%
    str_replace('historical plants species', 'historical plant species') %>%
    str_replace('heterogenous', 'heterogeneous') %>%
    str_replace('upward shift in', 'upward shifts in') %>%
    str_replace(' nd ', ' and ') %>%
    str_replace('altitiude', 'altitude') %>%
    ### truncate for ease of matching
    str_sub(1, 40)
}
ref_intsect <- refs_rubenstein %>%
  mutate(title = title_clean(title),
         first_author = author_clean(first_author)) %>%
  select(-doi, -author, -journal, -paperid) %>%
  full_join(refs_lenoir %>% 
              mutate(title = title_clean(title),
                     first_author = author_clean(first_author)),
            by = c('title', 'first_author')) %>%
  mutate(src = case_when(is.na(lenoir) ~ 'rubenstein',
                         is.na(rubenstein) ~ 'lenoir',
                         TRUE ~ 'both'))

table(ref_intsect$src)

x <- ref_intsect %>%
  filter(src != 'both') %>%
  arrange(first_author)

y <- x %>% janitor::get_dupes(first_author)
```
