---
title: "Data exploration: Rubenstein et al. 2023"
format: html
---

# Summary

Read in the various .csv files to see what they contain and how they relate to one another

# Data sources

* Rubenstein et al. (2023). Climate change and the global redistribution of biodiversity: Substantial variation in empirical support for expected range shifts. Environmental Evidence, 12(1), 7. https://doi.org/10.1186/s13750-023-00296-0
* CoRE (Contractions or Range Expansions) Database: Global Database of Species Range Shifts from 1802-2019: Data downloaded from https://www.sciencebase.gov/catalog/item/64147d6fd34eb496d1ceb497

# Methods

```{r}
library(tidyverse)
library(here)
```

## Read in .csvs individually

### Climate variables

```{r}
climate_df <- read_csv(here('Global Database of Species Range Shifts from 1802-2019', 'Climate.Variables.csv'))
head(climate_df)
```

```{r}
paperid_df <- read_csv(here('Global Database of Species Range Shifts from 1802-2019', 'PaperID.csv'))
names(paperid_df)
head(paperid_df)
```

```{r}
rshifts_df <- read_csv(here('Global Database of Species Range Shifts from 1802-2019', 'Range.Shifts.csv'))
names(rshifts_df)
```

